\documentclass[11pt,a4paper,twocolumn]{article}

% Fundamental Packets
\usepackage[utf8]{inputenc}         % chars codification
\usepackage[T1]{fontenc}            % font encoding
\usepackage[italian,english]{babel} % multilingual (english default)
\usepackage{amsmath,amssymb}        % maths symbols
\usepackage{graphicx}               % images
\usepackage{booktabs}               % professional tables
\usepackage{geometry}               % margin management
\usepackage{cite}                   % citations
\usepackage{url}                    % urls in bibliography

% margins
\geometry{left=1.5cm, right=1.5cm, top=1.5cm, bottom=2cm}

% Doc info
\title{Predicting Acute Oral Toxicity ($LD_{50}$): A Machine Learning and Deep Learning Approach using Molecular Descriptors and Graph Representations}
\author{Group 5 Members \\ 
        \small \textit{Abate Luca, Bernacchia Alessia, Pioda Tommaso, Villani Giacomo} \\
        \small \textit{Department of Innovative Technologies} \\
        \small \textit{SUPSI}}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Acute toxicity prediction is a cornerstone of drug discovery, as toxicity-related failures account for a significant portion of preclinical compound attrition. In this project, we utilize the Therapeutics Data Commons (TDC) $LD_{50}$ dataset, comprising 7,385 drug molecules, to develop a predictive framework. We implement a rigorous data pipeline involving chemical standardization and the extraction of hybrid descriptors, including Morgan, Avalon, and ErG fingerprints, alongside Dragon-derived physicochemical features. We evaluate the performance of classical machine learning models (Random Forest, Ridge, SVM, XGBoost) and deep learning architectures (CNN, GNN). Our methodology emphasizes data consistency and feature engineering to ensure that the models generalize effectively to novel chemical structures, ultimately aiming to reduce preclinical attrition and enhance drug safety profiles.
\end{abstract}

\section{Introduction}
The majority of drugs exhibit some degree of toxicity to human organisms. Accurate prediction of these effects is critical, as toxicity is one of the primary causes of compound attrition in the pharmaceutical pipeline. Studies indicate that approximately 70\% of all toxicity-related attrition occurs preclinically, yet these findings are strongly predictive of toxicities in humans. Consequently, early and accurate toxicity prediction can significantly reduce compound attribution and boost the likelihood of a drug successfully reaching the market.

This study focuses on the prediction of the median lethal dose ($LD_{50}$), a standard measure of acute toxicity. A significant challenge in this task is generalization: as drug structures evolve, models must maintain accuracy on novel drugs with low structural similarity to existing datasets. Our project addresses this by combining traditional physicochemical descriptors with advanced molecular fingerprints and graph-based deep learning, ensuring a comprehensive representation of chemical space.

\section{Materials \& Methods}

\subsection{Data Collection}
The dataset was obtained from the Therapeutics Data Commons (TDC) for the Acute Toxicity $LD_{50}$ task \cite{TDC_LD50}. The raw data provides Drug IDs, SMILES strings, and the target toxicity label ($Y$). The dataset consists of 7,385 small molecules, partitioned into the following sets:
\begin{itemize}
    \item \textbf{Training Set}: 5,169 samples (70\%)
    \item \textbf{Validation Set}: 738 samples (10\%)
    \item \textbf{Test Set}: 1,478 samples (20\%)
\end{itemize}

\subsection{Data Cleaning and Consistency}
Rigorous data cleaning was performed to ensure the quality of the inputs. The process involved:
\begin{itemize}
    \item \textbf{Standardization}: SMILES strings were checked for chemical validity using RDKit; 0 invalid strings were found across all sets.
    \item \textbf{Missing Values}: The dataset was confirmed to be complete, with no missing values in any columns.
    \item \textbf{Salt Stripping}: We utilized RDKit's \texttt{SaltRemover} to strip counter-ions and salts, ensuring that descriptors reflect only the parent organic molecule. No salts were detected in the current splits.
    \item \textbf{Duplicate and Inconsistency Handling}: No identical duplicates were found. However, we identified 15, 5, and 10 inconsistent SMILES in the train, valid, and test sets, respectively (where the same molecule was associated with different labels). We chose to retain these cases to represent natural assay variability, documenting them for evaluation.
\end{itemize}

\subsection{Molecular Descriptors and Feature Engineering}
We extracted three types of molecular fingerprints and two sets of physicochemical descriptors to capture diverse chemical properties without including the target $LD_{50}$ in the feature construction:
\begin{itemize}
    \item \textbf{Morgan Fingerprints}: Capture local atom environments via circular pathways.
    \item \textbf{Avalon Fingerprints}: Based on path-based enumeration and feature substructures \cite{Partex}.
    \item \textbf{ErG Fingerprints}: Capture pharmacophore features and topological distances \cite{Partex}.
    \item \textbf{General Descriptors}: Features like Molecular Weight, LogP, and TPSA \cite{Partex}.
    \item \textbf{Dragon Descriptors}: Advanced topological and electronic descriptors extracted based on established QSAR methodologies \cite{QSAR_Paper}.
\end{itemize}

\subsection{Model Selection and Embedded Feature Selection}
We implemented a range of models to compare traditional and deep learning approaches:
\begin{itemize}
    \item \textbf{Machine Learning}: Random Forest, Ridge Regression, Support Vector Machines (SVM), and XGBoost.
    \item \textbf{Deep Learning}: Convolutional Neural Networks (CNN) for sequence/grid data and Graph Neural Networks (GNN) to process molecular graphs directly.
    \item \textbf{Embedded Feature Selection}: To manage high dimensionality, we utilize embedded methods (e.g., Lasso or tree-based importance) that perform selection during training, identifying the most predictive features while reducing noise.
\end{itemize}

\section{Data Exploration and Processing}
% This section includes the details of the steps taken before modeling.
\subsection{Initial Statistics}
The initial dataset imported 7,385 drugs. Distribution analysis of the $LD_{50}$ values was conducted to understand the range of toxicity across the chemical space.

\subsection{Cleaning Results}
As noted in the Methods, the cleaning process verified the integrity of the TDC data. The lack of invalid SMILES and missing values suggests a high-quality initial collection. The removal of salts ensures that descriptors like Molecular Weight are not skewed by irrelevant counter-ions.

\subsection{Inconsistency Rationale}
The identified inconsistent samples (30 in total) were maintained to prevent the loss of experimental variability. In biological assays, $LD_{50}$ values can vary depending on conditions, and keeping these entries allows the models to potentially learn a more robust, averaged representation.

\subsection{Feature Integration}
By combining fingerprints (Avalon, Morgan, ErG) with Dragon descriptors, we create a hybrid feature space. These features are strictly structural and physicochemical, ensuring that no target leakage occurs during the learning process.

\section{Results}
Present your findings here. Use tables to summarize the data.

\begin{table}[h]
\centering
\caption{Dose-Response Data for Group 5 Substance.}
\begin{tabular}{ccc}
\toprule
Dose (mg/kg) & Mortality (\%) & Observations \\
\midrule
10  & 0   & No symptoms \\
50  & 20  & Lethargy \\
100 & 50  & $LD_{50}$ Threshold \\
500 & 100 & Rapid onset \\
\bottomrule
\end{tabular}
\end{table}

\section{Discussion \& Conclusion}
The results obtained for Group 5 indicate a toxicity class of [Class Number] according to the GHS (Globally Harmonized System). Comparing these results with historical data suggests that [Your Conclusion]. 

In conclusion, while $LD_{50}$ provides a fundamental baseline for toxicity, modern toxicology is increasingly moving towards alternative methods (3Rs: Replacement, Reduction, Refinement) to minimize animal testing while maintaining high safety standards.

\begin{thebibliography}{99}

\bibitem{TDC_LD50} 
Therapeutics Data Commons. \textit{Acute Toxicity LD50}. 
\url{https://tdcommons.ai/single_pred_tasks/tox/#acute-toxicity-ld50}

\bibitem{Partex} 
Partex ADMETrix. \textit{Advancing ADMET Prediction: A Hybrid Machine Learning Framework}. 
\url{https://github.com/partex-nv-opensource/tdc-submission/blob/main/src/ics-admet-prediction-tdc-versiontwo/report/Advancing%20ADMET%20Prediction_%20A%20Hybrid%20Machine%20Learning%20Framework%20(Partex%20ADMETrix).pdf}

\bibitem{QSAR_Paper} 
Zhu et al. \textit{Quantitative Structure-Activity Relationship Modeling of Rat Acute Toxicity by Oral Exposure}. 
\url{https://pubs.acs.org/doi/abs/10.1021/tx900189p}

\end{thebibliography}

\end{document}